---
title: "Using External Dosage Tables with Medusa"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using External Dosage Tables with Medusa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

Medusa now supports two genotype ingestion paths:

1. OMOP Genomic CDM via `VARIANT_OCCURRENCE`
2. External dosage extracts via `formatExternalDosageTable()`

The external-dosage path is intended for resources such as All of Us,
UK Biobank, PLINK exports, and Hail-derived dosage tables where the EHR
phenotypes live in OMOP but genotypes are managed outside the OMOP
Genomic CDM.

# Step 1: Prepare the instrument table

```{r}
library(Medusa)

instruments <- createInstrumentTable(
  snpId = c("rs2228145", "rs4129267"),
  effectAllele = c("C", "T"),
  otherAllele = c("A", "C"),
  betaZX = c(0.35, 0.31),
  seZX = c(0.015, 0.016),
  pvalZX = c(1e-100, 1e-80),
  eaf = c(0.39, 0.37),
  geneRegion = rep("IL6R", 2)
)
```

# Step 2: Standardize the external dosage extract

The formatter accepts wide or long data. In practice, this table would come
from a PLINK/PGEN/VCF/Hail extraction keyed on the same rs IDs as the study
instrument table.

```{r}
externalDosage <- data.frame(
  person_id = c(1, 2, 3),
  rs2228145 = c(0, 1, 2),
  rs4129267 = c(2, 1, 0),
  stringsAsFactors = FALSE
)

formattedDosage <- formatExternalDosageTable(
  dosageTable = externalDosage,
  instrumentTable = instruments
)

formattedDosage$dosageTable
```

# Step 3: Build the OMOP cohort while reusing the external dosages

`buildMRCohort()` still uses OMOP CDM for outcomes, demographics, and
eligibility. The only difference is `genotypeSource = "externalDosageTable"`.

```{r eval = FALSE}
cohortData <- buildMRCohort(
  connectionDetails = connectionDetails,
  cdmDatabaseSchema = "cdm",
  cohortDatabaseSchema = "results",
  cohortTable = "cohort",
  outcomeCohortId = 1234,
  instrumentTable = instruments,
  genotypeSource = "externalDosageTable",
  externalDosageTable = formattedDosage
)
```

The returned cohort retains a harmonized instrument table as an attribute so
downstream modeling uses the same allele coding that was applied during cohort
construction.

```{r}
names(attributes(formattedDosage))
```

# Step 4: Fit the Medusa model as usual

Once the cohort has been built, the rest of the Medusa pipeline is unchanged.

```{r eval = FALSE}
profile <- fitOutcomeModel(
  cohortData = cohortData,
  instrumentTable = instruments,
  siteId = "all_of_us_replication"
)
```

# All of Us workflow notes

For All of Us specifically, the intended workflow is:

1. Define the outcome cohort in the OMOP-based Researcher Workbench.
2. Extract the instrument SNP dosages from the Controlled Tier genomic assets.
3. Convert that extract into a Medusa dosage table with
   `formatExternalDosageTable()`.
4. Use `buildMRCohort(..., genotypeSource = "externalDosageTable")` to merge
   the OMOP cohort with the external genotype matrix.
5. Run the standard Medusa diagnostics, cis-MR, colocalization, and reporting
   workflow.

This path avoids requiring the OMOP Genomic CDM extension at All of Us sites
while preserving the same downstream Medusa interfaces.
