---
title: "IL-6 Signaling and Colorectal Cancer: A Complete MR Walkthrough"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{IL-6 Signaling and Colorectal Cancer: A Complete MR Walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

## Scientific Background

Interleukin-6 (IL-6) is a pleiotropic cytokine involved in inflammation, immune
regulation, and hematopoiesis. Elevated IL-6 levels have been associated with
colorectal cancer risk in observational studies, but these associations may be
confounded by lifestyle factors, adiposity, or reverse causation (cancer causing
inflammation rather than vice versa). The IL6R missense variant rs2228145
(Asp358Ala) is a biologically well-characterized perturbation of the pathway:
it increases soluble IL-6 receptor levels and alters downstream signaling
(Ferreira et al., 2013).

Mendelian Randomization can address this question: **Does genetically proxied
IL-6 signaling causally affect colorectal cancer risk?**

The IL-6 receptor (IL6R) gene region on chromosome 1q21 contains well-characterized
variants (notably rs2228145/Asp358Ala) that alter IL-6 signaling. These serve as
strong genetic instruments for MR analysis. Recent MR work on circulating
interleukins and colorectal cancer has not established a clear causal signal for
IL-6/IL-6 receptor traits, so this vignette should be read as a realistic
workflow example anchored in a real scientific question, not as a claim that the
synthetic estimate below reproduces a published effect size ("Circulating
interleukins and risk of colorectal cancer: a Mendelian randomization study",
2023).

## Two-Sample MR in One Page

This vignette uses **two-sample Mendelian randomization (MR)**. In plain terms:

1. We obtain the SNP-exposure associations (`beta_ZX`) from an external GWAS.
   Here, the "exposure" is genetically proxied IL-6 signaling.

2. We estimate the SNP-outcome associations (`beta_ZY`) in a separate outcome
   dataset. In Medusa, that outcome dataset can be a single OMOP site or a
   federated network of sites.

3. For each instrument, the core causal signal is the **Wald ratio**:
   `beta_ZY / beta_ZX`. If multiple SNPs are used, Medusa's primary analysis
   combines them into a fixed weighted allele score and applies the same logic
   to that score.

This is called "two-sample" because the exposure and outcome associations come
from two distinct samples, not because two hospitals are required. In this
package, the exposure side typically comes from a public GWAS, while the outcome
side comes from local OMOP-linked genotype data.

The usual MR assumptions are:

- **Relevance**: the selected SNPs truly predict the exposure of interest.
- **Independence**: the SNPs are not associated with major confounders of the
  exposure-outcome relationship.
- **Exclusion restriction**: the SNPs affect the outcome primarily through the
  exposure pathway being studied.

The rest of the vignette shows how Medusa operationalizes that design in a
federated setting: the coordinator provides `beta_ZX`, sites estimate `beta_ZY`,
and only profile-likelihood summaries leave each site.

## Step 1: Instrument Assembly

In a real analysis, you would query OpenGWAS:

```{r instruments-real, eval=FALSE}
library(Medusa)

# Query OpenGWAS for IL-6 receptor GWAS
instruments <- getMRInstruments(
  exposureTraitId = "ieu-a-1119",  # IL-6 receptor levels
  pThreshold = 5e-8,
  r2Threshold = 0.001,
  kb = 10000,
  ancestryPopulation = "EUR"
)
```

For this vignette, we use real IL6R-region SNP identifiers but fixed,
deterministic summary-statistic values so the document builds offline. In a real
analysis, the exact `beta_ZX` and `se_ZX` values should come from the external
GWAS used to define the score.

```{r instruments-simulated}
library(Medusa)

# These are literature-grounded IL6R variants. The summary-statistic values
# below are illustrative placeholders for a reproducible offline vignette.
instruments <- createInstrumentTable(
  snpId = c("rs2228145", "rs4129267", "rs7529229", "rs4845625",
            "rs6689306", "rs12118721", "rs4453032"),
  effectAllele = c("C", "T", "T", "T", "G", "T", "A"),
  otherAllele = c("A", "C", "C", "C", "A", "C", "G"),
  betaZX = c(0.35, 0.31, 0.28, 0.22, 0.18, 0.15, 0.12),
  seZX = c(0.015, 0.016, 0.017, 0.018, 0.020, 0.022, 0.025),
  pvalZX = c(1e-100, 1e-80, 1e-60, 1e-30, 1e-18, 1e-10, 1e-6),
  eaf = c(0.39, 0.37, 0.35, 0.42, 0.28, 0.31, 0.22),
  geneRegion = rep("IL6R", 7)
)

cat(sprintf("Assembled %d instruments from the IL6R region.\n", nrow(instruments)))
cat(sprintf("F-statistics range: %.0f to %.0f\n",
            min(instruments$fStatistic), max(instruments$fStatistic)))
```

## Step 2: Outcome Cohort Definition

In OMOP CDM (with the [OMOP Genomic CDM](https://github.com/OHDSI/Genomic-CDM)
extension), incident colorectal cancer would typically be defined using:

- **SNOMED concepts**: Malignant neoplasm of colon (concept ID 4089661),
  Malignant neoplasm of rectum (concept ID 4180790)
- **ICD-10-CM**: C18.x (colon), C19 (rectosigmoid junction), C20 (rectum)
- **Exclusion criteria**: Prior history of any cancer, inflammatory bowel disease
- **Washout period**: >= 365 days of prior observation

```{r cohort-definition, eval=FALSE}
# This would run at each OMOP CDM site
cohort <- buildMRCohort(
  connectionDetails = connectionDetails,
  cdmDatabaseSchema = "cdm",
  cohortDatabaseSchema = "results",
  cohortTable = "cohort",
  outcomeCohortId = 1234,  # Your colorectal cancer cohort ID
  instrumentTable = instruments,
  genomicDatabaseSchema = "genomics",  # Schema with VARIANT_OCCURRENCE
  washoutPeriod = 365,
  excludePriorOutcome = TRUE
)
```

## Step 3: Simulated Analysis

To keep the vignette fully executable, we simulate a local site dataset while
preserving the same IL6R score definition used in the instrument table above.
This avoids a common MR mistake: fitting the outcome model with one score and
then computing the Wald ratio with a different denominator.

```{r simulate-and-fit}
# Simulate one site using the same allele-score definition that Medusa fits
set.seed(2026)
n <- 8000

snpMatrix <- sapply(instruments$eaf, function(maf) rbinom(n, 2, maf))
colnames(snpMatrix) <- paste0("snp_", gsub("[^a-zA-Z0-9]", "_", instruments$snp_id))

scoreWeights <- instruments$beta_ZX / (instruments$se_ZX^2)
scoreWeights <- scoreWeights / sum(abs(scoreWeights))
scoreBetaZX <- sum(scoreWeights * instruments$beta_ZX)

confounder1 <- rnorm(n)
confounder2 <- rbinom(n, 1, 0.5)

exposure <- scoreBetaZX * as.numeric(snpMatrix %*% scoreWeights) +
  0.3 * confounder1 + 0.3 * confounder2 + rnorm(n)

trueEffect <- -0.16
logOdds <- trueEffect * exposure + 0.3 * confounder1 + 0.3 * confounder2
outcome <- rbinom(n, 1, 1 / (1 + exp(-logOdds)))

cohortData <- data.frame(
  person_id = seq_len(n),
  outcome = outcome,
  snpMatrix,
  confounder_1 = confounder1,
  confounder_2 = confounder2,
  stringsAsFactors = FALSE
)

betaGrid <- seq(-2, 2, by = 0.02)

# Fit the site-specific outcome model
profile <- fitOutcomeModel(
  cohortData = cohortData,
  covariateData = NULL,
  instrumentTable = instruments,
  betaGrid = betaGrid,
  siteId = "site_A"
)
```

## Checking Two-Sample MR Assumptions with Medusa

Medusa does not "prove" the MR assumptions, but it gives you a structured way
to look for violations before trusting the causal estimate.

- **Relevance**: `runInstrumentDiagnostics()` reports per-SNP F-statistics. Weak
  instruments (`F < 10`) can bias MR estimates.
- **Independence**: the instrument PheWAS checks whether SNPs are associated
  with observed covariates. Strong associations with likely confounders can be a
  warning sign.
- **Exclusion restriction**: Medusa cannot verify this directly, but unexpected
  PheWAS hits, allele-frequency mismatches, or major missingness can indicate
  pleiotropy, data-quality problems, or harmonization issues that threaten this
  assumption.

The package also has a `negativeControlResults` slot in the diagnostics object,
but this vignette focuses on the currently executable checks: F-statistics,
covariate associations, allele-frequency comparison, and genotype missingness.

```{r assumption-diagnostics}
# Build a diagnostics-ready copy of the synthetic cohort
diagnosticCohort <- cohortData
diagnosticCohort$personId <- diagnosticCohort$person_id
diagnosticCohort$exposure <- exposure

# Add a small amount of missingness so the missingness report is visible
set.seed(2027)
missingIdx <- sample.int(n, size = round(0.12 * n))
diagnosticCohort$snp_rs4453032[missingIdx] <- NA_integer_

# Create synthetic covariates and add one deliberately SNP-correlated trait
diagnosticCovariates <- simulateCovariateData(
  n = n,
  nCovariates = 8,
  seed = 2026
)
names(diagnosticCovariates)[names(diagnosticCovariates) == "person_id"] <- "personId"
diagnosticCovariates$proxy_inflammation_trait <-
  as.numeric(scale(diagnosticCohort$snp_rs2228145 + rnorm(n, sd = 0.05)))

diagnostics <- suppressWarnings(
  runInstrumentDiagnostics(
    cohortData = diagnosticCohort,
    covariateData = diagnosticCovariates,
    instrumentTable = instruments,
    exposureProxyConceptIds = 1L,
    pValueThreshold = 1e-4
  )
)

# Assumption 1: relevance
print(diagnostics$fStatistics[, c("snp_id", "fStatistic", "source", "weakFlag")])

# Assumptions 2 and 3: look for suspicious covariate associations
subset(
  diagnostics$phewasResults,
  significant,
  select = c("snp_id", "covariate_name", "pval")
)

# Data integrity checks that support harmonisation review
subset(
  diagnostics$afComparison,
  select = c("snp_id", "eaf_gwas", "eaf_cohort", "eaf_diff", "discrepancyFlag")
)
subset(
  diagnostics$missingnessReport,
  select = c("snp_id", "pct_missing", "highMissingFlag")
)

# High-level summary of which checks need follow-up
diagnostics$diagnosticFlags
```

In practice, you would use these diagnostics as a screening step:

- Proceed with more confidence when instruments are strong, PheWAS signals are
  limited, and allele frequencies look consistent.
- Investigate or prune instruments when specific SNPs are weak, associated with
  likely confounders, or show allele-frequency discrepancies.
- Treat missingness and harmonization anomalies as data-quality issues to fix
  before interpreting the MR result.

## Step 4: Federated Pooling

To illustrate federated pooling without requiring three live data partners, we
retain the observed `site_A` profile and create two additional site profiles
with the same score definition but slightly different information content.

```{r pool-profiles}
makeSyntheticSite <- function(baseProfile, siteId, infoScale, betaShift, caseScale) {
  peak <- baseProfile$betaHat + betaShift
  info <- (1 / (baseProfile$seHat^2)) * infoScale
  logLik <- -0.5 * info * (baseProfile$betaGrid - peak)^2
  logLik <- logLik - max(logLik)

  synthetic <- list(
    siteId = siteId,
    betaGrid = baseProfile$betaGrid,
    logLikProfile = logLik,
    nCases = max(50L, as.integer(round(baseProfile$nCases * caseScale))),
    nControls = max(50L, as.integer(round(baseProfile$nControls * caseScale))),
    snpIds = baseProfile$snpIds,
    diagnosticFlags = baseProfile$diagnosticFlags,
    betaHat = peak,
    seHat = sqrt(1 / info),
    scoreDefinition = baseProfile$scoreDefinition
  )
  class(synthetic) <- "medusaSiteProfile"
  synthetic
}

siteProfiles <- list(
  site_A = profile,
  site_B = makeSyntheticSite(profile, "site_B", infoScale = 0.9,
                             betaShift = 0.01, caseScale = 0.85),
  site_C = makeSyntheticSite(profile, "site_C", infoScale = 1.1,
                             betaShift = -0.01, caseScale = 1.15)
)

# Pool
combined <- poolLikelihoodProfiles(siteProfiles)

# Visualize
plotLikelihoodProfile(
  combinedProfile = combined,
  siteProfileList = siteProfiles,
  title = "Profile Likelihood: IL-6 Signaling on Colorectal Cancer"
)
```

## Step 5: MR Estimate

```{r compute-estimate}
estimate <- computeMREstimate(combined, instruments)

cat(sprintf("Causal OR for IL-6 signaling on colorectal cancer: %.3f\n",
            estimate$oddsRatio))
cat(sprintf("95%% CI: [%.3f, %.3f]\n", estimate$orCiLower, estimate$orCiUpper))
cat(sprintf("P-value: %.2e\n", estimate$pValue))
```

## Step 6: Sensitivity Analyses

Because the one-shot pooled likelihood is defined on the shared allele score,
pleiotropy-robust summarized-data methods are a secondary analysis. Here we use
per-SNP estimates from the same synthetic site. Medusa's internal Steiger
implementation does not support binary outcomes; for binary-outcome Steiger
filtering, use `engine = "TwoSampleMR"`, which delegates to
`TwoSampleMR::steiger_filtering()`. We force Medusa's internal engine here so
the vignette remains fully reproducible without requiring `TwoSampleMR`.

```{r sensitivity}
# Fit the same site in per-SNP mode for summarized-data sensitivity methods
profilePerSnp <- fitOutcomeModel(
  cohortData = cohortData,
  covariateData = NULL,
  instrumentTable = instruments,
  betaGrid = betaGrid,
  analysisType = "perSNP",
  siteId = "site_A"
)

sensitivity <- runSensitivityAnalyses(
  profilePerSnp$perSnpEstimates,
  methods = c("IVW", "MREgger", "WeightedMedian", "LeaveOneOut"),
  engine = "internal"
)

# Summary
print(sensitivity$summary)
```

## Interpretation for Drug Development

The MR analysis provides genetic evidence regarding the causal role of IL-6
signaling in colorectal cancer risk, but the numerical result in this vignette
comes from synthetic outcome data. The worked example is therefore about correct
analysis mechanics, not about claiming a real protective or harmful effect.

In a real target-validation study, the primary interpretation would rest on
four checks:

1. **Instrument fidelity**: The same allele score must define the local outcome
   model and the external `beta_ZX` denominator.

2. **Pleiotropy robustness**: Concordance across IVW, MR-Egger, weighted
   median, and leave-one-out analyses is more informative than a single point
   estimate.

3. **Clinical triangulation**: Any MR signal should be compared against human
   genetics, disease biology, and trial safety data for IL-6 pathway blockade.

4. **Literature context**: The current colorectal-cancer MR literature for
   circulating interleukins remains mixed, so this is best treated as a
   motivating oncology use case rather than a settled causal claim.

For readers newer to MR, the key conceptual takeaway is that the package is not
estimating "the effect of carrying a risk SNP." It is using inherited genetic
variation as a proxy for a modifiable biology (here, IL-6 signaling), then
asking whether people with genetically shifted pathway activity also differ in
colorectal cancer risk in the expected direction.

## References

- [Burgess, S., Butterworth, A., & Thompson, S. G. (2013). Mendelian
  randomization analysis with multiple genetic variants using summarized data.
  *Genetic Epidemiology*, 37(7), 658-665.](https://pmc.ncbi.nlm.nih.gov/articles/PMC4377079/)

- [Bowden, J., Davey Smith, G., & Burgess, S. (2015). Mendelian randomization
  with invalid instruments: effect estimation and bias detection through Egger
  regression. *International Journal of Epidemiology*, 44(2), 512-525.](https://pubmed.ncbi.nlm.nih.gov/26050253/)

- [Bowden, J., et al. (2016). Consistent estimation in Mendelian randomization
  with some invalid instruments using a weighted median estimator. *Genetic
  Epidemiology*, 40(4), 304-314.](https://pmc.ncbi.nlm.nih.gov/articles/PMC4849733/)

- [Ferreira, R. C., Freitag, D. F., Cutler, A. J., Howson, J. M. M., et al.
  (2013). Functional IL6R 358Ala allele impairs classical IL-6 receptor
  signaling and influences risk of diverse inflammatory diseases. *PLoS
  Genetics*, 9(4), e1003444.](https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1003444)

- [Circulating interleukins and risk of colorectal cancer: a Mendelian
  randomization study. (2023). *Journal of Cancer Research and Clinical
  Oncology*. PMID: 37525405.](https://pubmed.ncbi.nlm.nih.gov/37525405/)
