---
title: "IL-6 Signaling and Colorectal Cancer: A Complete MR Walkthrough"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{IL-6 Signaling and Colorectal Cancer: A Complete MR Walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

## Scientific Background

Interleukin-6 (IL-6) is a pleiotropic cytokine involved in inflammation, immune
regulation, and hematopoiesis. Elevated IL-6 levels have been associated with
colorectal cancer risk in observational studies, but these associations may be
confounded by lifestyle factors, adiposity, or reverse causation (cancer causing
inflammation rather than vice versa).

Mendelian Randomization can address this question: **Does genetically proxied
IL-6 signaling causally affect colorectal cancer risk?**

The IL-6 receptor (IL6R) gene region on chromosome 1q21 contains well-characterized
variants (notably rs2228145/Asp358Ala) that alter IL-6 signaling. These serve as
strong genetic instruments for MR analysis.

## Step 1: Instrument Assembly

In a real analysis, you would query OpenGWAS:

```{r instruments-real, eval=FALSE}
library(Medusa)

# Query OpenGWAS for IL-6 receptor GWAS
instruments <- getMRInstruments(
  exposureTraitId = "ieu-a-1119",  # IL-6 receptor levels
  pThreshold = 5e-8,
  r2Threshold = 0.001,
  kb = 10000,
  ancestryPopulation = "EUR"
)
```

For this vignette, we create instruments from known IL6R-region SNPs:

```{r instruments-simulated}
library(Medusa)

instruments <- createInstrumentTable(
  snpId = c("rs2228145", "rs4129267", "rs7529229", "rs4845625",
            "rs6689306", "rs12118721", "rs4453032"),
  effectAllele = c("C", "T", "T", "T", "G", "T", "A"),
  otherAllele = c("A", "C", "C", "C", "A", "C", "G"),
  betaZX = c(0.35, 0.31, 0.28, 0.22, 0.18, 0.15, 0.12),
  seZX = c(0.015, 0.016, 0.017, 0.018, 0.020, 0.022, 0.025),
  pvalZX = c(1e-100, 1e-80, 1e-60, 1e-30, 1e-18, 1e-10, 1e-6),
  eaf = c(0.39, 0.37, 0.35, 0.42, 0.28, 0.31, 0.22),
  geneRegion = rep("IL6R", 7)
)

cat(sprintf("Assembled %d instruments from the IL6R region.\n", nrow(instruments)))
cat(sprintf("F-statistics range: %.0f to %.0f\n",
            min(instruments$fStatistic), max(instruments$fStatistic)))
```

## Step 2: Outcome Cohort Definition

In OMOP CDM, incident colorectal cancer would typically be defined using:

- **SNOMED concepts**: Malignant neoplasm of colon (concept ID 4089661),
  Malignant neoplasm of rectum (concept ID 4180790)
- **ICD-10-CM**: C18.x (colon), C19 (rectosigmoid junction), C20 (rectum)
- **Exclusion criteria**: Prior history of any cancer, inflammatory bowel disease
- **Washout period**: >= 365 days of prior observation

```{r cohort-definition, eval=FALSE}
# This would run at each OMOP CDM site
cohort <- buildMRCohort(
  connectionDetails = connectionDetails,
  cdmDatabaseSchema = "cdm",
  cohortDatabaseSchema = "results",
  cohortTable = "cohort",
  outcomeCohortId = 1234,  # Your colorectal cancer cohort ID
  instrumentTable = instruments,
  genomicLinkageSchema = "genomics",
  genomicLinkageTable = "genotype_data",
  washoutPeriod = 365,
  excludePriorOutcome = TRUE
)
```

## Step 3: Simulated Analysis

For this vignette, we use simulated data to demonstrate the full pipeline:

```{r simulate-and-fit}
# Simulate data with a modest protective effect (OR ~ 0.85, beta ~ -0.16)
simData <- simulateMRData(
  n = 8000,
  nSnps = 7,
  trueEffect = -0.16,
  seed = 2026
)

# Fit outcome model
profile <- fitOutcomeModel(
  cohortData = simData$data,
  covariateData = NULL,
  instrumentTable = simData$instrumentTable,
  betaGrid = seq(-2, 2, by = 0.02),
  siteId = "site_vignette"
)
```

## Step 4: Federated Pooling

Simulating three sites pooling their results:

```{r pool-profiles}
# Simulate profiles from 3 sites
siteProfiles <- simulateSiteProfiles(
  nSites = 3,
  betaGrid = seq(-2, 2, by = 0.02),
  trueBeta = -0.16,
  nPerSite = 5000,
  seed = 2026
)

# Pool
combined <- poolLikelihoodProfiles(siteProfiles)

# Visualize
plotLikelihoodProfile(
  combinedProfile = combined,
  siteProfileList = siteProfiles,
  title = "Profile Likelihood: IL-6 Signaling on Colorectal Cancer"
)
```

## Step 5: MR Estimate

```{r compute-estimate}
estimate <- computeMREstimate(combined, instruments)

cat(sprintf("Causal OR for IL-6 signaling on colorectal cancer: %.3f\n",
            estimate$oddsRatio))
cat(sprintf("95%% CI: [%.3f, %.3f]\n", estimate$orCiLower, estimate$orCiUpper))
cat(sprintf("P-value: %.2e\n", estimate$pValue))
```

## Step 6: Sensitivity Analyses

```{r sensitivity}
# Create per-SNP estimates (simulated)
set.seed(2026)
nSnps <- 7
betaZX <- instruments$beta_ZX
perSnp <- data.frame(
  snp_id = instruments$snp_id,
  beta_ZY = -0.16 * betaZX + rnorm(nSnps, 0, 0.01),
  se_ZY = runif(nSnps, 0.015, 0.030),
  beta_ZX = betaZX,
  se_ZX = instruments$se_ZX
)

sensitivity <- runSensitivityAnalyses(perSnp)

# Summary
print(sensitivity$summary)

# Forest plot
plotSensitivityForest(sensitivity)
```

## Interpretation for Drug Development

The MR analysis provides genetic evidence regarding the causal role of IL-6
signaling in colorectal cancer risk. Key considerations for translating MR
findings to drug development:

1. **Direction of effect**: A protective OR < 1 for genetically increased IL-6
   signaling would suggest that IL-6 *inhibition* could increase cancer risk,
   relevant for anti-IL-6 drug safety monitoring.

2. **Sensitivity analysis concordance**: If IVW, MR-Egger, and weighted median
   estimates agree in direction and significance, this strengthens the causal
   interpretation.

3. **MR-Egger intercept**: A non-significant intercept suggests no evidence of
   directional pleiotropy, supporting the validity of the MR assumptions.

4. **Comparison to clinical trials**: MR estimates represent lifelong genetic
   differences, while drug effects are time-limited. The direction should agree,
   but magnitudes may differ.
