---
title: "Course 6: Doing and Critiquing a Two-Sample MR Study"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Course 6: Doing and Critiquing a Two-Sample MR Study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

## What you'll learn

By the end of this chapter, you should be able to:

- describe an end-to-end two-sample MR study workflow,
- identify major design and interpretation checkpoints,
- use Medusa outputs to critique an analysis,
- explain when an MR estimate should be treated cautiously.

## Why this matters

By this point, you know the logic of MR and the mechanics of estimation. The
last step is judgment: can you design a credible study, read one critically,
and know when the output supports a cautious causal interpretation?

## A practical study checklist

Before running an MR study, ask:

1. Is the exposure well defined and biologically coherent?
2. Are the instruments strong enough?
3. Are the outcome data appropriate for the question?
4. Is allele harmonization trustworthy?
5. Are the sensitivity analyses informative?
6. Are the limitations stated honestly?

This checklist is often more important than the final p-value.

## A small end-to-end synthetic analysis

```{r setup-analysis}
library(Medusa)

set.seed(1006)
beta_grid <- seq(-1.5, 1.5, by = 0.05)

sim_data <- simulateMRData(
  n = 5000,
  nSnps = 6,
  trueEffect = 0.30,
  seed = 1006
)

profile <- fitOutcomeModel(
  cohortData = sim_data$data,
  covariateData = NULL,
  instrumentTable = sim_data$instrumentTable,
  betaGrid = beta_grid,
  siteId = "site_A",
  analysisType = "perSNP"
)

combined <- poolLikelihoodProfiles(list(site_A = profile))
estimate <- computeMREstimate(combined, sim_data$instrumentTable)
sensitivity <- runSensitivityAnalyses(
  profile$perSnpEstimates,
  methods = c("IVW", "MREgger", "WeightedMedian", "LeaveOneOut"),
  engine = "internal"
)
```

### Main estimate

```{r main-estimate}
estimate$betaMR
estimate$ciLower
estimate$ciUpper
estimate$oddsRatio
```

### Sensitivity summary

```{r sensitivity-summary}
sensitivity$summary
```

## How to critique the analysis

### 1. Instrument quality

Check:

- approximate F-statistics,
- whether instruments come from a biologically sensible region,
- whether the number of SNPs is enough to support meaningful sensitivity
  analyses.

### 2. Harmonization and data integrity

Check:

- allele labels,
- palindromic SNP handling,
- allele-frequency consistency,
- genotype missingness.

### 3. Estimator agreement

Check:

- whether IVW, weighted median, and MR-Egger point in the same general
  direction,
- whether one SNP dominates leave-one-out results,
- whether uncertainty is very large.

### 4. Interpretation discipline

Even a well-conducted MR result should be described carefully.

Good language:

- "consistent with a causal effect,"
- "supports a likely causal role,"
- "evidence is compatible with..."

Overstated language:

- "proves causation,"
- "settles the question,"
- "guarantees a drug will work."

**Key takeaway:** the output of MR is evidence, not certainty.

## Common reasons not to trust an MR estimate

- weak instruments,
- obvious horizontal pleiotropy,
- severe disagreement across methods,
- poor harmonization,
- implausible biology,
- strong chance of sample selection artifacts,
- outcome definitions that do not match the scientific question.

## Exercise set 1

### Exercise 1

If IVW is positive, weighted median is near zero, and leave-one-out changes sign
when one SNP is removed, what should you conclude?

<details class="solution">
<summary>Show solution</summary>

You should conclude that the result is unstable and should not be treated as
robust causal evidence without deeper investigation. One SNP may be overly
influential, or the methods may be reacting to invalid instruments.

</details>

### Exercise 2

Why is "MR proves causation" usually too strong?

<details class="solution">
<summary>Show solution</summary>

Because MR depends on assumptions that can fail, especially exclusion
restriction and independence. It can strengthen causal inference substantially,
but it does not remove uncertainty completely.

</details>

### Exercise 3

Name two things you would inspect before trusting a surprising MR result.

<details class="solution">
<summary>Show solution</summary>

Reasonable answers include instrument strength, allele harmonization,
population-structure concerns, sensitivity-analysis agreement, leave-one-out
results, and biological plausibility.

</details>

## Exercise set 2: critique a mock abstract

Mock claim:

> We found a statistically significant MR estimate linking biomarker X to
> disease Y, therefore biomarker X is definitively causal and should be
> targeted therapeutically.

What is missing from this claim?

<details class="solution">
<summary>Show solution</summary>

The claim ignores instrument validity, pleiotropy, sensitivity analyses,
harmonization quality, uncertainty, and the difference between causal evidence
and therapeutic success. A responsible abstract would discuss the assumptions
and supporting diagnostics rather than making a definitive statement.

</details>

## Where to go next

You now have the conceptual foundation to:

- read ordinary two-sample MR papers critically,
- understand why sensitivity analyses matter,
- understand how Medusa's federated workflow connects to textbook MR logic.

For next steps inside the package:

- [Getting Started with Medusa](GettingStarted.html) for a runnable workflow,
- [IL-6 Signaling and Colorectal Cancer](ColorectalCancerIL6Example.html) for a
  scientific walkthrough,
- [Federated Analysis Guide](FederatedAnalysisGuide.html) for network-level
  deployment.

## Chapter summary

- A good MR study is judged by design quality and robustness, not only by the
  headline estimate.
- You should inspect instrument strength, harmonization, diagnostics, and
  sensitivity analyses together.
- Medusa provides tools that support this critical workflow, but the final
  judgment remains scientific and epidemiologic.

## References

- Davies, N.M., Holmes, M.V., & Davey Smith, G. (2018). Reading Mendelian
  randomisation studies: a guide for clinicians. *BMJ*, 362, k601.
- Burgess, S., Davey Smith, G., Davies, N.M., et al. (2019). Guidelines for
  performing Mendelian randomization investigations. *Wellcome Open Research*,
  4, 186.
- Skrivankova, V.W., Richmond, R.C., Woolf, B.A.R., et al. (2021).
  Strengthening the reporting of observational studies in epidemiology using
  Mendelian randomization: the STROBE-MR statement. *JAMA*, 326(16),
  1614--1621.

## End of course

You have now moved from observational causal problems, to instrumental
variables, to two-sample MR mechanics, to sensitivity analyses, and finally to
Medusa's federated implementation and critical interpretation.
