---
title: "Course 3: The Mechanics of Two-Sample MR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Course 3: The Mechanics of Two-Sample MR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

## What you'll learn

By the end of this chapter, you should be able to:

- explain what "two-sample" means,
- interpret `beta_ZX` and `beta_ZY`,
- compute a Wald ratio by hand,
- explain why allele harmonization matters,
- understand how multiple-SNP summarized-data MR is organized.

## Why this matters

This chapter is where MR becomes operational. The main conceptual move in
two-sample MR is simple:

- one dataset tells us how strongly each SNP predicts the exposure,
- another dataset tells us how strongly the same SNP predicts the outcome.

Those two pieces are combined into a causal estimate.

## What does "two-sample" mean?

In **two-sample MR**, the SNP-exposure and SNP-outcome associations are estimated
in different samples.

Typical pattern:

- exposure sample: a large external GWAS,
- outcome sample: your cohort, biobank, or federated network.

This does **not** mean that two hospitals are required. It means the exposure
and outcome associations come from distinct datasets.

## The two key quantities

For a given SNP \(Z\):

- \(\beta_{ZX}\): association of SNP \(Z\) with exposure \(X\),
- \(\beta_{ZY}\): association of SNP \(Z\) with outcome \(Y\).

If the SNP affects the outcome only through the exposure, then a simple causal
estimate is:

\[
\beta_{MR} = \frac{\beta_{ZY}}{\beta_{ZX}}
\]

This is the **Wald ratio**.

### Interpreting the ratio

If:

- a SNP increases LDL-C by 0.20 units, and
- the same SNP increases log-odds of CAD by 0.10,

then the Wald ratio is \(0.10 / 0.20 = 0.50\). That means:

- the implied causal effect is 0.50 log-odds units of CAD per 1-unit increase
  in LDL-C.

## One-SNP worked example

```{r one-snp}
beta_zx <- 0.20
beta_zy <- 0.10
beta_mr <- beta_zy / beta_zx
beta_mr
```

That is the most basic MR estimate possible.

## Why the SNP signs must match

Suppose one dataset defines the effect allele as allele `A`, but the other
defines the effect allele as allele `G` at the same SNP. Then the reported
effect directions can point in opposite directions.

If you do not harmonize them first, the ratio can be wrong even when both input
studies were internally correct.

### Mini harmonization example

```{r harmonization-table}
harmonization_example <- data.frame(
  snp_id = c("rs1", "rs2", "rs3"),
  effect_allele = c("A", "C", "T"),
  other_allele = c("G", "T", "C"),
  beta_ZX = c(0.20, -0.10, 0.15),
  beta_ZY = c(0.08, -0.06, 0.09),
  stringsAsFactors = FALSE
)

harmonization_example
```

If one SNP is coded with the opposite effect allele in one dataset, both
`beta_ZX` and `beta_ZY` for that SNP must be flipped together before the ratio
is computed.

**Key takeaway:** harmonization is not a cosmetic step. It is necessary to keep
the numerator and denominator talking about the same allele.

## Multiple SNPs

Most MR analyses use more than one SNP.

For each SNP \(j\), we have:

\[
\beta_{MR,j} = \frac{\beta_{ZY,j}}{\beta_{ZX,j}}
\]

Those SNP-specific ratios can then be combined, often using inverse-variance
weighting.

Here is a simple summarized-data example:

```{r multiple-snps}
summary_data <- data.frame(
  snp_id = c("rsA", "rsB", "rsC"),
  beta_ZX = c(0.20, 0.12, 0.30),
  se_ZX = c(0.04, 0.03, 0.06),
  beta_ZY = c(0.10, 0.05, 0.17),
  se_ZY = c(0.03, 0.02, 0.05),
  effect_allele = c("A", "C", "T"),
  other_allele = c("G", "T", "C"),
  eaf = c(0.30, 0.45, 0.25),
  stringsAsFactors = FALSE
)

summary_data$wald_ratio <- summary_data$beta_ZY / summary_data$beta_ZX
summary_data
```

## Why Medusa is a little different

Textbook summarized-data MR often combines SNP-specific ratio estimates
directly.

Medusa supports those methods in `runSensitivityAnalyses()`, but its **primary
federated estimator** is different:

- sites estimate a pooled SNP-outcome signal through a weighted allele score,
- they share a profile log-likelihood rather than person-level data,
- the coordinator reconstructs the MR estimate from the pooled profile and the
  score's SNP-exposure association.

So Medusa uses standard summarized-data MR ideas for sensitivity analysis, but a
profile-likelihood route for the main federated estimate.

## Exercise set 1

### Exercise 1

If `beta_ZX = 0.25` and `beta_ZY = 0.05`, what is the Wald ratio?

<details class="solution">
<summary>Show solution</summary>

The Wald ratio is `0.05 / 0.25 = 0.20`.

</details>

### Exercise 2

Why must `beta_ZX` and `beta_ZY` be aligned to the same effect allele before
forming a ratio?

<details class="solution">
<summary>Show solution</summary>

Because the ratio only makes sense when the numerator and denominator describe
the effect of the same allele. If one association is for allele `A` and the
other is for allele `G`, the directions can be inconsistent and the ratio can
be wrong.

</details>

### Exercise 3

In one sentence, what does "two-sample" refer to in two-sample MR?

<details class="solution">
<summary>Show solution</summary>

It means the SNP-exposure associations and SNP-outcome associations are
estimated in different samples or datasets.

</details>

## Exercise set 2: small coding practice

Use the small table below to compute SNP-specific Wald ratios in R.

```{r coding-practice}
practice_df <- data.frame(
  snp = c("rs1", "rs2"),
  beta_ZX = c(0.10, 0.25),
  beta_ZY = c(0.03, 0.09)
)

practice_df$beta_MR <- practice_df$beta_ZY / practice_df$beta_ZX
practice_df
```

<details class="solution">
<summary>Show solution</summary>

The key code is:

```r
practice_df$beta_MR <- practice_df$beta_ZY / practice_df$beta_ZX
practice_df
```

The resulting ratio estimates are 0.30 and 0.36. They are similar but not
identical, which is typical in multi-SNP MR because each SNP-specific estimate
contains sampling noise and may also be affected by different biases.

</details>

## Chapter summary

- Two-sample MR combines SNP-exposure and SNP-outcome associations from
  different datasets.
- `beta_ZX` describes SNP to exposure association.
- `beta_ZY` describes SNP to outcome association.
- The simplest causal estimate is the Wald ratio `beta_ZY / beta_ZX`.
- Allele harmonization is essential.
- Medusa uses conventional summarized-data logic for sensitivity analyses, but
  a federated profile-likelihood approach for its main pooled estimate.

## Next chapter

Next: [Course 4: Estimation and sensitivity analysis](Course04-EstimationAndSensitivityAnalysis.html)

If you want to see the package workflow in action before moving on, see
[Getting Started with Medusa](GettingStarted.html).
