---
title: "Course 5: From Theory to the Medusa Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Course 5: From Theory to the Medusa Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

## What you'll learn

By the end of this chapter, you should be able to:

- map textbook two-sample MR quantities onto Medusa objects,
- explain the roles of the coordinator and local sites,
- explain why Medusa shares profile likelihoods instead of person-level data,
- walk through the main Medusa analysis functions on simulated data.

## Why this matters

The earlier chapters explained two-sample MR as a summarized-data method. Medusa
implements that same causal logic in a privacy-preserving federated workflow.

The exposure side still comes from external summary statistics. The main
difference is how the outcome side is estimated and shared.

## The Medusa division of labor

### Coordinator tasks

The coordinator typically:

- defines the scientific question,
- assembles or imports the instrument table,
- distributes the instrument table to participating sites,
- receives site-level summary artifacts,
- pools them and computes the final MR estimate.

### Site tasks

Each site typically:

- extracts the outcome cohort,
- harmonizes genotype coding,
- fits the outcome model locally,
- exports a profile likelihood summary,
- optionally exports per-SNP summary estimates for sensitivity analyses.

**Key takeaway:** Medusa separates the exposure side, the local outcome-model
fitting, and the final pooled MR estimation.

## How Medusa represents the MR ingredients

In textbook notation:

- `beta_ZX`: SNP-exposure associations from external GWAS,
- `beta_ZY`: SNP-outcome association estimated from the outcome sample.

In Medusa:

- the instrument table stores the SNP-exposure side (`beta_ZX`, `se_ZX`,
  alleles, effect allele frequency),
- `fitOutcomeModel()` estimates the outcome-side signal locally,
- `poolLikelihoodProfiles()` combines site summaries,
- `computeMREstimate()` converts the pooled profile into the final MR estimate.

## Why profile likelihood is shared

Medusa's primary federated estimate does not ask each site to export all
person-level outcomes or raw genotypes. Instead, each site evaluates a profile
log-likelihood over a grid of candidate SNP-outcome effect values.

The coordinator receives:

- the grid,
- the log-likelihood values,
- metadata such as case counts,
- the allele-score definition.

Because log-likelihoods from independent sites add, the coordinator can pool
the profiles exactly by summation.

This is what makes the approach both federated and statistically principled.

## A runnable Medusa example

```{r simulate-data}
library(Medusa)

beta_grid <- seq(-2, 2, by = 0.05)

sim_data <- simulateMRData(
  n = 4000,
  nSnps = 8,
  trueEffect = 0.40,
  seed = 1005
)

head(sim_data$instrumentTable[, c("snp_id", "beta_ZX", "se_ZX", "eaf")])
```

### Step 1: fit the local outcome model

```{r fit-site}
profile_site_a <- fitOutcomeModel(
  cohortData = sim_data$data,
  covariateData = NULL,
  instrumentTable = sim_data$instrumentTable,
  betaGrid = beta_grid,
  siteId = "site_A"
)

names(profile_site_a)
```

This local profile is Medusa's site-level representation of the SNP-outcome
evidence.

### Step 2: create partner site summaries

To simulate federation, we create two compatible partner profiles that reuse the
same allele-score definition.

```{r partner-sites}
make_partner_profile <- function(template, site_id, shift, info_scale, n_cases, n_controls) {
  partner <- template
  partner$siteId <- site_id
  partner$betaHat <- template$betaHat + shift
  partner$seHat <- template$seHat / sqrt(info_scale)
  partner$logLikProfile <- -0.5 * info_scale *
    ((template$betaGrid - partner$betaHat) / template$seHat)^2
  partner$logLikProfile <- partner$logLikProfile - max(partner$logLikProfile)
  partner$nCases <- n_cases
  partner$nControls <- n_controls
  partner
}

site_profiles <- list(
  site_A = profile_site_a,
  site_B = make_partner_profile(profile_site_a, "site_B", shift = 0.04,
                                info_scale = 1.10, n_cases = 230, n_controls = 1770),
  site_C = make_partner_profile(profile_site_a, "site_C", shift = -0.03,
                                info_scale = 0.95, n_cases = 210, n_controls = 1790)
)
```

### Step 3: pool the profiles

```{r pool-profiles}
combined_profile <- poolLikelihoodProfiles(site_profiles)

combined_profile$nSites
combined_profile$totalCases
combined_profile$totalControls
```

### Step 4: recover the MR estimate

```{r mr-estimate}
mr_estimate <- computeMREstimate(combined_profile, sim_data$instrumentTable)

mr_estimate$betaMR
mr_estimate$ciLower
mr_estimate$ciUpper
mr_estimate$oddsRatio
```

## Mapping objects back to theory

- `sim_data$instrumentTable` corresponds to the SNP-exposure side.
- `profile_site_a$betaHat` is the local site's fitted score-to-outcome effect.
- `combined_profile$logLikProfile` is the pooled evidence about the outcome-side
  parameter.
- `mr_estimate$betaMR` is the causal estimate after dividing the pooled
  SNP-outcome signal by the exposure effect of the same fitted score.

This last point is important: Medusa's main federated estimator is built around
the exact allele score used locally, not a generic average SNP effect.

**Key takeaway:** Medusa preserves the logic of two-sample MR while changing the
way outcome information is communicated across sites.

## Optional per-SNP mode for sensitivity analyses

```{r per-snp}
profile_per_snp <- fitOutcomeModel(
  cohortData = sim_data$data,
  covariateData = NULL,
  instrumentTable = sim_data$instrumentTable,
  betaGrid = beta_grid,
  siteId = "site_A",
  analysisType = "perSNP"
)

head(profile_per_snp$perSnpEstimates)
```

Those `perSnpEstimates` are the package's bridge back to standard summarized-data
MR sensitivity analyses.

## Exercise set 1

### Exercise 1

In Medusa, where does the SNP-exposure information live?

<details class="solution">
<summary>Show solution</summary>

It lives in the instrument table, which carries quantities such as `beta_ZX`,
`se_ZX`, allele labels, and effect allele frequencies.

</details>

### Exercise 2

Why does Medusa pool log-likelihood profiles instead of person-level records?

<details class="solution">
<summary>Show solution</summary>

Because the site-level log-likelihood profile is a privacy-preserving summary
that can be added across independent sites to recover the pooled evidence
without sharing person-level genotypes or outcomes.

</details>

### Exercise 3

What does `analysisType = "perSNP"` add beyond the main federated estimate?

<details class="solution">
<summary>Show solution</summary>

It provides per-SNP outcome associations that can be used for summarized-data
sensitivity analyses such as IVW, MR-Egger, weighted median, and leave-one-out.

</details>

## Exercise set 2: inspect Medusa objects

Use the objects created above and answer:

1. Which object stores the number of pooled sites?
2. Which object stores the final odds ratio?

<details class="solution">
<summary>Show solution</summary>

`combined_profile$nSites` stores the number of pooled sites, and
`mr_estimate$oddsRatio` stores the final odds ratio.

</details>

## Chapter summary

- Medusa keeps the exposure side in the instrument table and estimates the
  outcome side locally.
- Sites share profile-likelihood summaries rather than person-level records.
- The coordinator pools those profiles and computes the MR estimate from the
  same fitted allele score used at the sites.
- Per-SNP output is available when you want standard summarized-data
  sensitivity analyses.

## References

- Suchard, M.A., Simpson, S.E., Zorych, I., Ryan, P., & Madigan, D. (2013).
  Massive parallelization of serial inference algorithms for a complex
  generalized linear model. *ACM Transactions on Modeling and Computer
  Simulation*, 23(1), 1--17.
- Duan, R., Boland, M.R., Liu, Z., et al. (2020). Learning from electronic
  health records across multiple sites: a communication-efficient and
  privacy-preserving distributed algorithm. *Journal of the American Medical
  Informatics Association*, 27(3), 376--385.
- Hemani, G., Zheng, J., Elsworth, B., et al. (2018). The MR-Base platform
  supports systematic causal inference across the human phenome. *eLife*, 7,
  e34408.

## Next chapter

Next: [Course 6: Doing and critiquing a two-sample MR study](Course06-DoingAndCritiquingATwoSampleMRStudy.html)

For more package detail after this chapter, see
[Getting Started with Medusa](GettingStarted.html) and
[Federated Analysis Guide](FederatedAnalysisGuide.html).
