---
title: "Federated Analysis Guide for Network Coordinators"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Federated Analysis Guide for Network Coordinators}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

This guide walks OHDSI network coordinators through setting up and running a
federated Medusa analysis across multiple sites. The key principle: **only
profile log-likelihood vectors leave each site** — no individual-level data
is shared.

The code chunks in this vignette are templates, not executable examples. They
are shown with `eval = FALSE` because every network needs to substitute its own
connection details, cohort IDs, file paths, and governance controls.

## What Data Leaves Each Site

| Shared with coordinator | NOT shared |
|-------------------------|------------|
| Log-likelihood profile (numeric vector, ~600 numbers) | Individual genotypes |
| Number of cases and controls | Person-level outcomes |
| Optional diagnostic flag summary (logical values) | Covariate values |
| Optional per-SNP summary estimates for sensitivity analyses | Raw SNP-by-person data |
| Site identifier | Demographics |

The profile vector is a smooth curve that represents the aggregate statistical
evidence at a site. It cannot be reverse-engineered to identify individuals.
The required export already includes model-level flags such as low case count or
an MLE at the grid boundary.
If a network wants IVW, MR-Egger, or weighted-median sensitivity analyses, each
site can also share one row per SNP containing `beta_ZY` and `se_ZY`. Those are
still aggregate summaries, but they are optional and separate from the main
pooled likelihood workflow.

## Site Setup Requirements

### OMOP CDM Requirements

- OMOP CDM version 5.3 or 5.4
- Standard tables: PERSON, OBSERVATION_PERIOD, CONDITION_OCCURRENCE
- A cohort table with the outcome cohort pre-defined (e.g., via ATLAS)

### Genomic Linkage Table

Each site needs a table linking person_id to SNP genotypes:

```sql
CREATE TABLE genomics.genotype_data (
  person_id BIGINT NOT NULL,
  snp_id VARCHAR(50) NOT NULL,
  genotype INTEGER NOT NULL  -- 0, 1, or 2 (count of effect alleles)
);
```

If the person identifier column has a different name (e.g., `subject_id`),
specify it via the `genomicPersonIdColumn` parameter.

### R Package Dependencies

```{r install-site}
# Required at each site
remotes::install_github("OHDSI/Medusa")

# This installs: DatabaseConnector, SqlRender, Cyclops, FeatureExtraction
```

### Ancestry Principal Components (Optional but Recommended)

A table with ancestry PCs controls for population stratification:

```sql
CREATE TABLE genomics.ancestry_pcs (
  person_id BIGINT NOT NULL,
  pc1 FLOAT, pc2 FLOAT, pc3 FLOAT, pc4 FLOAT, pc5 FLOAT,
  pc6 FLOAT, pc7 FLOAT, pc8 FLOAT, pc9 FLOAT, pc10 FLOAT
);
```

## Template Site Analysis Script

Send this script to each participating site, customized with their local
connection details:

```{r site-script}
# ============================================================
# Medusa Site Analysis Script
# Study: [YOUR STUDY NAME]
# Site: [SITE NAME]
# Date: [DATE]
# ============================================================

library(Medusa)

# ----- Site-specific configuration -----
connectionDetails <- DatabaseConnector::createConnectionDetails(
  dbms = "postgresql",              # Change to your DBMS
  server = "localhost/ohdsi",       # Change to your server
  user = "ohdsi_user",             # Change to your user
  password = keyring::key_get("ohdsi_db")  # Use secure credential storage
)

cdmDatabaseSchema <- "cdm"           # Change to your CDM schema
cohortDatabaseSchema <- "results"     # Change to your results schema
cohortTable <- "cohort"               # Change if different
outcomeCohortId <- 1234               # Provided by coordinator
genomicLinkageSchema <- "genomics"    # Change to your genomics schema
genomicLinkageTable <- "genotype_data"
siteId <- "site_A"                    # Unique identifier for this site

# ----- Load instrument table (provided by coordinator) -----
instrumentTable <- read.csv("instruments.csv", stringsAsFactors = FALSE)

# ----- Step 1: Build cohort -----
cohortData <- buildMRCohort(
  connectionDetails = connectionDetails,
  cdmDatabaseSchema = cdmDatabaseSchema,
  cohortDatabaseSchema = cohortDatabaseSchema,
  cohortTable = cohortTable,
  outcomeCohortId = outcomeCohortId,
  instrumentTable = instrumentTable,
  genomicLinkageSchema = genomicLinkageSchema,
  genomicLinkageTable = genomicLinkageTable,
  washoutPeriod = 365,
  excludePriorOutcome = TRUE
)

# ----- Step 2: Build covariates -----
covariateData <- buildMRCovariates(
  connectionDetails = connectionDetails,
  cdmDatabaseSchema = cdmDatabaseSchema,
  cohortDatabaseSchema = cohortDatabaseSchema,
  cohortTable = cohortTable,
  outcomeCohortId = outcomeCohortId
)

# ----- Step 3: Run diagnostics -----
diagnostics <- runInstrumentDiagnostics(
  cohortData = cohortData,
  covariateData = covariateData,
  instrumentTable = instrumentTable
)

# ----- Step 4: Fit outcome model -----
# Use the SAME betaGrid as all other sites
betaGrid <- seq(-3, 3, by = 0.01)

profile <- fitOutcomeModel(
  cohortData = cohortData,
  covariateData = covariateData,
  instrumentTable = instrumentTable,
  betaGrid = betaGrid,
  siteId = siteId,
  analysisType = "alleleScore"
)

# ----- Step 5: Export and share -----
# Required for the primary federated MR estimate:
exportSiteProfile(profile, outputDir = ".", prefix = "medusa")

# Optional: share a one-line-per-check summary of the richer diagnostics object.
utils::write.csv(
  data.frame(
    check = names(diagnostics$diagnosticFlags),
    flag = unname(diagnostics$diagnosticFlags),
    stringsAsFactors = FALSE
  ),
  sprintf("medusa_diagnostic_flags_%s.csv", siteId),
  row.names = FALSE
)

# Optional: only run this block if the coordinator requested per-SNP summaries
# for IVW / MR-Egger / weighted-median sensitivity analyses.
profilePerSnp <- fitOutcomeModel(
  cohortData = cohortData,
  covariateData = covariateData,
  instrumentTable = instrumentTable,
  betaGrid = betaGrid,
  siteId = siteId,
  analysisType = "perSNP"
)
utils::write.csv(
  profilePerSnp$perSnpEstimates,
  sprintf("medusa_per_snp_%s.csv", siteId),
  row.names = FALSE
)

message("Analysis complete. Share the required profile CSVs and, if requested, the per-SNP summary CSV.")
```

## Coordinator Pooling Script

After collecting profile CSV files from all sites:

```{r coordinator-script}
library(Medusa)

# Load instrument table
instrumentTable <- read.csv("instruments.csv", stringsAsFactors = FALSE)

# Import site profiles from CSV files
siteProfiles <- list(
  site_A = importSiteProfile("medusa_profile_site_A.csv"),
  site_B = importSiteProfile("medusa_profile_site_B.csv"),
  site_C = importSiteProfile("medusa_profile_site_C.csv")
)

# Pool
combined <- poolLikelihoodProfiles(siteProfiles)

# Estimate
estimate <- computeMREstimate(combined, instrumentTable)

# Optional: collect simple diagnostic flag summaries if sites shared them.
diagnosticFlagPaths <- c(
  "medusa_diagnostic_flags_site_A.csv",
  "medusa_diagnostic_flags_site_B.csv",
  "medusa_diagnostic_flags_site_C.csv"
)
if (all(file.exists(diagnosticFlagPaths))) {
  siteDiagnosticFlags <- lapply(diagnosticFlagPaths, read.csv, stringsAsFactors = FALSE)
}

# Optional: set this to NULL unless per-SNP CSVs were collected.
sensitivity <- NULL

# Optional: fixed-effect pool the per-SNP summaries before sensitivity analyses.
# Each site's per-SNP CSV should contain one row per SNP with beta_ZY and se_ZY.
perSnpPaths <- c(
  site_A = "medusa_per_snp_site_A.csv",
  site_B = "medusa_per_snp_site_B.csv",
  site_C = "medusa_per_snp_site_C.csv"
)
if (all(file.exists(perSnpPaths))) {
  perSnpBySite <- lapply(perSnpPaths, read.csv, stringsAsFactors = FALSE)

  stackedPerSnp <- do.call(rbind, perSnpBySite)
  splitPerSnp <- split(stackedPerSnp, stackedPerSnp$snp_id)

  perSnpPooled <- do.call(
    rbind,
    lapply(splitPerSnp, function(df) {
      weights <- 1 / (df$se_ZY^2)
      data.frame(
        snp_id = df$snp_id[1],
        effect_allele = df$effect_allele[1],
        other_allele = df$other_allele[1],
        eaf = df$eaf[1],
        beta_ZY = sum(weights * df$beta_ZY) / sum(weights),
        se_ZY = sqrt(1 / sum(weights)),
        beta_ZX = df$beta_ZX[1],
        se_ZX = df$se_ZX[1],
        pval_ZX = df$pval_ZX[1],
        stringsAsFactors = FALSE
      )
    })
  )

  sensitivity <- runSensitivityAnalyses(
    perSnpPooled,
    methods = c("IVW", "MREgger", "WeightedMedian", "LeaveOneOut"),
    engine = "internal"
  )
}

# Report
generateMRReport(
  mrEstimate = estimate,
  sensitivityResults = sensitivity,
  combinedProfile = combined,
  siteProfileList = siteProfiles,
  instrumentTable = instrumentTable,
  exposureLabel = "IL-6 signaling",
  outcomeLabel = "Colorectal cancer"
)
```

## Secure File Transfer

Profile CSV files should be transferred securely between sites and coordinator.
Options include:

- SFTP with encrypted credentials
- Institutional secure file sharing (e.g., Box, SharePoint with encryption)
- OHDSI network file transfer protocols (if available)

The required profile CSV files are human-readable and typically very small
(< 100 KB) because they contain only numeric grid values and summary
statistics. Optional per-SNP summary CSVs are also compact. Using CSV ensures
that every value leaving a site can be inspected and audited before transfer.

## Coordinator Pre-flight Checklist

Before pooling, verify:

- Every site used the same `instruments.csv` file.
- Every site used the same `betaGrid` values.
- Every imported profile has the expected site identifier and non-zero case count.
- If per-SNP summaries were requested, each file uses the same SNP IDs and allele coding.

## Handling Different OMOP Versions

Sites running OMOP CDM v5.3 vs v5.4 can participate together. The SQL templates
in Medusa use only core CDM tables that are consistent across versions. If a site
uses non-standard table names, these can be configured via function parameters.

## Troubleshooting

### "No persons have genotype data"

- Verify the genomic linkage table exists and has data
- Check that `person_id` column names match (use `genomicPersonIdColumn` parameter)
- Ensure the cohort and genotype table share person_id values

### "Profile likelihood is flat"

- Instruments may be too weak at this site
- Check F-statistics in diagnostics
- Verify that genotype data is coded correctly (0/1/2)

### "MLE is at grid boundary"

- Expand the `betaGrid` range (e.g., `seq(-5, 5, by = 0.01)`)
- Ensure the same `betaGrid` is used at all sites

### Weak instrument warning (F < 10)

- The instrument explains very little outcome variance at this site
- Consider excluding weak instruments or using more instruments
- Results may be biased toward the null
