---
title: "Getting Started with Medusa"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with Medusa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

## What is Medusa?

Medusa (**M**endelian **E**stimation in **D**istributed **S**tandardized
**A**nalytics) is an R package that implements two-sample Mendelian
Randomization (MR) within the OHDSI ecosystem using the OMOP Common Data Model.

**Mendelian Randomization** uses genetic variants as instrumental variables
to estimate causal effects of exposures on outcomes. Because genetic variants
are randomly allocated at conception (like a natural randomized trial), MR can
provide evidence for causality from observational data.

**Two-sample MR** separates the analysis into two parts:

1. **SNP-exposure associations** (beta_ZX): Obtained from published GWAS
   (typically from large biobank studies)
2. **SNP-outcome associations** (beta_ZY): Estimated from your own cohort data

The causal estimate is the **Wald ratio**: beta_MR = beta_ZY / beta_ZX.

This vignette is a runnable sandbox: every example below uses synthetic data,
so the goal is to show the workflow and the shape of the outputs rather than
to make a scientific claim.

## Installation

```{r install, eval=FALSE}
# Install from GitHub
remotes::install_github("OHDSI/Medusa")

# This will also install OHDSI dependencies:
# DatabaseConnector, SqlRender, Cyclops, FeatureExtraction
```

## The Federated Approach

Medusa's key innovation is **one-shot federated pooling**. Here's how it works:

1. The **coordinator** assembles genetic instruments from public GWAS data
2. Each **site** (hospital/biobank with OMOP CDM + genomic data):
   - Extracts the outcome cohort and genotype data locally
   - Fits a regression model locally
   - Computes a **profile log-likelihood curve** (a vector of numbers)
   - Shares ONLY this vector — no individual-level data leaves the site
3. The **coordinator** sums the log-likelihood vectors across sites
4. The causal estimate is extracted from the combined curve

This is privacy-preserving by design: the shared data is a smooth numeric vector
that cannot be reverse-engineered to identify individual patients.

## Quick Start with Simulated Data

No database connection needed for this example.

```{r quickstart}
library(Medusa)

betaGrid <- seq(-2, 2, by = 0.05)

# Simulate MR data with a known causal effect of 0.5 (log-OR)
simData <- simulateMRData(
  n = 5000,
  nSnps = 10,
  trueEffect = 0.5,
  seed = 42
)

# Look at the instrument table
head(simData$instrumentTable[, c("snp_id", "beta_ZX", "se_ZX", "eaf")])
```

### Fit the Outcome Model

```{r fit-model}
# Fit the outcome model and get the profile likelihood
profileSiteA <- fitOutcomeModel(
  cohortData = simData$data,
  covariateData = NULL,
  instrumentTable = simData$instrumentTable,
  betaGrid = betaGrid,
  siteId = "site_A"
)

# The profile contains no individual-level data
names(profileSiteA)
cat(sprintf("beta_ZY estimate: %.3f (SE: %.3f)\n",
            profileSiteA$betaHat, profileSiteA$seHat))
```

### Pool Profiles (Simulating Federation)

```{r pool}
# Create synthetic partner sites that reuse the same allele-score definition.
# This keeps the pooled likelihood compatible with the denominator used later
# in computeMREstimate().
makePartnerProfile <- function(template, siteId, shift, infoScale, nCases, nControls) {
  partner <- template
  partner$siteId <- siteId
  partner$betaHat <- template$betaHat + shift
  partner$seHat <- template$seHat / sqrt(infoScale)
  partner$logLikProfile <- -0.5 * infoScale *
    ((template$betaGrid - partner$betaHat) / template$seHat)^2
  partner$logLikProfile <- partner$logLikProfile - max(partner$logLikProfile)
  partner$nCases <- nCases
  partner$nControls <- nControls
  partner
}

profiles <- list(
  site_A = profileSiteA,
  site_B = makePartnerProfile(profileSiteA, "site_B", shift = 0.06,
                              infoScale = 1.1, nCases = 295, nControls = 2705),
  site_C = makePartnerProfile(profileSiteA, "site_C", shift = -0.04,
                              infoScale = 0.9, nCases = 260, nControls = 2740)
)

# Pool the profiles
combined <- poolLikelihoodProfiles(profiles)
cat(sprintf("Pooled %d sites: %d total cases, %d total controls\n",
            combined$nSites, combined$totalCases, combined$totalControls))
```

### Compute the MR Estimate

```{r estimate}
estimate <- computeMREstimate(combined, simData$instrumentTable)

cat(sprintf("Causal estimate (beta_MR): %.3f\n", estimate$betaMR))
cat(sprintf("95%% CI: [%.3f, %.3f]\n", estimate$ciLower, estimate$ciUpper))
cat(sprintf("P-value: %.2e\n", estimate$pValue))
cat(sprintf("Odds ratio: %.3f\n", estimate$oddsRatio))
```

The key detail is that the pooled profile carries the exact allele-score
definition used at each site. Passing the same instrument table back into
`computeMREstimate()` keeps the Wald ratio denominator aligned with the fitted
score rather than with an unrelated simulated instrument set.

### Run Sensitivity Analyses

```{r sensitivity}
# Create per-SNP estimates for sensitivity analyses
set.seed(42)
nSnps <- 10
instrumentSummary <- simulateInstrumentTable(nSnps = nSnps, seed = 42)
perSnp <- data.frame(
  snp_id = instrumentSummary$snp_id,
  beta_ZY = 0.5 * instrumentSummary$beta_ZX + rnorm(nSnps, 0, 0.02),
  se_ZY = rep(0.02, nSnps),
  beta_ZX = instrumentSummary$beta_ZX,
  se_ZX = instrumentSummary$se_ZX,
  effect_allele = instrumentSummary$effect_allele,
  other_allele = instrumentSummary$other_allele,
  eaf = instrumentSummary$eaf
)

results <- runSensitivityAnalyses(
  perSnp,
  methods = c("IVW", "MREgger", "WeightedMedian", "LeaveOneOut"),
  engine = "internal"
)
print(results$summary)
```

## Understanding the Report

The `generateMRReport()` function creates a self-contained HTML report with:

1. **Executive Summary** — Plain-language interpretation with strength rating
2. **Instrument Summary** — Table of all SNPs with F-statistics
3. **Likelihood Profile** — Visual of site and combined likelihood curves
4. **Main Result** — Odds ratio with confidence interval
5. **Sensitivity Analyses** — Forest plot comparing methods
6. **Diagnostics** — instrument PheWAS, allele-frequency checks, missingness
7. **Methods Section** — Auto-generated text for manuscripts

## FAQ

**What genomic data linkage is required?**
Your OMOP CDM site needs a genomic linkage table mapping person_id to SNP
genotypes (coded as 0/1/2 for allele count). This is increasingly common
in biobank-linked health systems.

**What if my site has no genomic data?**
Medusa requires at least one site with genomic data linked to OMOP CDM.
Sites without genomic data cannot participate in the federated analysis.

**How many sites are needed?**
A single site is sufficient. More sites increase statistical power and
allow for cross-site validation. The federated approach adds value with
2+ sites.

**What OMOP CDM version is required?**
Medusa supports OMOP CDM v5.3 and v5.4.
