---
title: "Negative Control Outcomes and Empirical Calibration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Negative Control Outcomes and Empirical Calibration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  fig.width = 7,
  fig.height = 5
)
if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(path = ".", quiet = TRUE, export_all = FALSE, helpers = FALSE)
}
simulateMRData <- getFromNamespace("simulateMRData", "Medusa")
simulateNegativeControlOutcomes <- getFromNamespace(
  "simulateNegativeControlOutcomes",
  "Medusa"
)
runNegativeControlAnalysis <- getFromNamespace(
  "runNegativeControlAnalysis",
  "Medusa"
)
```

## Why Negative Controls?

A two-sample Mendelian Randomization study can produce a statistically
significant result even when there is no true causal effect. This happens
when one of the core MR assumptions is violated --- typically the exclusion
restriction (instruments affect the outcome only through the exposure) or
independence (instruments are not associated with confounders). Standard
sensitivity analyses like MR-Egger and weighted median can flag certain
patterns of pleiotropy, but they rely on parametric assumptions of their
own.

**Negative control outcomes** provide an empirical, assumption-light check.
The idea is simple: pick a set of outcomes that you are confident the
exposure does *not* cause. Run the full MR pipeline on each one. If the
analysis is well-behaved, the resulting effect estimates should be centered
at zero. If they are not, the pipeline has a systematic problem ---
residual pleiotropy, population stratification, or some other source of
bias --- and the primary result should be interpreted with caution.

This is the same logic behind the OHDSI empirical calibration framework
used in large-scale observational studies (Schuemie et al., 2014; 2018),
adapted here for Mendelian Randomization.


## What Makes a Good Negative Control?

A negative control outcome should satisfy two criteria:

1. **No causal effect of the exposure.** You must be confident, based on
   biology and prior evidence, that the exposure cannot plausibly cause the
   outcome. For example, if you are studying the effect of LDL cholesterol
   on coronary artery disease, a negative control might be an accidental
   fracture --- there is no biological mechanism linking LDL to fracture
   risk.

2. **Susceptible to the same biases as the primary analysis.** The negative
   control should be measured in the same population, using the same data
   pipeline, and subject to the same potential confounders. An outcome that
   is immune to confounding would not detect confounding-driven bias in
   your primary analysis.

### Practical guidance for selecting negative controls

- **Use published lists.** The OHDSI community maintains curated negative
  control outcome lists for common drug-outcome studies. These are a good
  starting point.
- **Aim for 20--100 negative controls.** More is better for fitting the
  systematic error distribution. Fewer than 10 makes calibration unstable.
- **Mix prevalences.** Include both common and rare outcomes to stress-test
  the pipeline across different case counts.
- **Avoid outcomes on the causal pathway.** If LDL lowers cardiovascular
  risk, then outcomes downstream of cardiovascular disease (e.g., heart
  failure hospitalization) are not valid negative controls.
- **Avoid outcomes caused by the instruments.** If a genetic variant in the
  *HMGCR* region also affects liver enzymes through a pathway independent
  of LDL, then liver enzyme elevation is not a valid negative control ---
  it would produce a non-null MR estimate even without bias.


## Using Negative Controls in Medusa

### Option 1: Database extraction with `buildMRCohort()`

If your negative control outcomes are defined as cohorts in the OMOP CDM
(e.g., from ATLAS), pass their cohort definition IDs when building the
MR cohort:
```{r buildMRCohort-nc, eval = FALSE}
cohortData <- buildMRCohort(
  connectionDetails = connectionDetails,
  cdmDatabaseSchema = "cdm",
  cohortDatabaseSchema = "results",
  cohortTable = "cohort",
  outcomeCohortId = 100,
  instrumentTable = instruments,
  negativeControlCohortIds = c(201, 202, 203, 204, 205,
                               206, 207, 208, 209, 210)
)
```

This adds binary columns `nc_outcome_201`, `nc_outcome_202`, etc., to
the returned data frame. Each column is 1 if the person had an entry in
that cohort, 0 otherwise.

### Option 2: Simulated data for testing

For demonstrations and unit tests, Medusa provides a simulation helper
that adds synthetic negative control outcome columns to existing cohort
data. These outcomes are generated independently of genotype, so the true
effect is exactly zero.

```{r simulate-nc}
simData <- simulateMRData(n = 2000, nSnps = 5, seed = 42)
cohortData <- simulateNegativeControlOutcomes(
  simData$data,
  nControls = 25,
  seed = 123
)

# The original columns are preserved, plus 25 nc_outcome_* columns
ncCols <- grep("^nc_outcome_", names(cohortData), value = TRUE)
cat(length(ncCols), "negative control outcomes added\n")
head(ncCols)
```


### Running the negative control analysis

`runNegativeControlAnalysis()` takes the cohort data and instrument table,
fits the MR pipeline on each negative control outcome, and returns a
summary of the results:

```{r run-nc}
ncResults <- suppressMessages(
  runNegativeControlAnalysis(
    cohortData = cohortData,
    instrumentTable = simData$instrumentTable
  )
)

str(ncResults, max.level = 1)
```

The returned object has class `medusaNegativeControls` and contains four
elements:

- **`ncEstimates`**: A data frame with one row per negative control
  outcome, containing the MR effect estimate (`beta_MR`), standard error
  (`se_MR`), and p-value.
- **`calibration`**: The fitted systematic error model from the
  `EmpiricalCalibration` package (NULL if the package is not installed).
- **`calibratedPrimary`**: Calibrated p-value and confidence interval for
  the primary estimate (NULL if no primary estimate was supplied).
- **`biasDetected`**: A logical flag. TRUE if the distribution of negative
  control estimates is significantly different from zero (one-sample
  t-test, p < 0.05).


## Interpreting the Results

### Step 1: Are the negative control estimates centered at zero?

The most basic check. Look at the distribution of `beta_MR` across
negative control outcomes:

```{r nc-distribution}
estimates <- ncResults$ncEstimates

cat("Number of NC outcomes analyzed:", nrow(estimates), "\n")
cat("Mean beta_MR:", round(mean(estimates$beta_MR, na.rm = TRUE), 4), "\n")
cat("SD of beta_MR:", round(sd(estimates$beta_MR, na.rm = TRUE), 4), "\n")
cat("Bias detected:", ncResults$biasDetected, "\n")
```

If `biasDetected` is FALSE, the pipeline is behaving as expected on
outcomes with no true effect. If TRUE, there is evidence of systematic
bias --- proceed with caution when interpreting the primary result.

### Step 2: Visualize the null distribution

A histogram of the negative control MR estimates should be roughly
centered at zero:

```{r nc-histogram, fig.cap = "Distribution of negative control MR effect estimates. Centered at zero indicates no systematic bias."}
hist(
  estimates$beta_MR,
  breaks = 15,
  main = "Negative Control MR Estimates",
  xlab = expression(hat(beta)[MR]),
  col = "#264553",
  border = "white"
)
abline(v = 0, col = "#E69528", lwd = 2, lty = 2)
```

### Step 3: Examine individual negative control p-values

Under the null, p-values should be uniformly distributed. An excess of
small p-values suggests systematic bias:

```{r nc-pvalues}
# How many NC outcomes have p < 0.05?
nSig <- sum(estimates$pval < 0.05, na.rm = TRUE)
nTotal <- sum(!is.na(estimates$pval))
cat(sprintf(
  "%d of %d NC outcomes significant at p < 0.05 (expected: ~%.0f)\n",
  nSig, nTotal, 0.05 * nTotal
))
```


### Step 4: Empirical calibration of the primary estimate

If the `EmpiricalCalibration` package is installed, Medusa fits a
systematic error model to the negative control distribution. This model
captures the mean and variance of the systematic error, which can then be
used to adjust the primary estimate's p-value and confidence interval.

To use this feature, pass your primary estimate when running the analysis:

```{r nc-calibration, eval = FALSE}
ncResults <- runNegativeControlAnalysis(
  cohortData = cohortData,
  instrumentTable = instruments,
  primaryEstimate = list(
    betaMR  = 0.30,
    seMR    = 0.10,
    pValue  = 0.003
  )
)

# The calibrated primary estimate accounts for systematic error
ncResults$calibratedPrimary
# $calibratedP        -- p-value after accounting for systematic error
# $calibratedCiLower  -- adjusted 95% CI lower bound
# $calibratedCiUpper  -- adjusted 95% CI upper bound
```

The calibrated p-value is typically larger (more conservative) than the
nominal p-value because it accounts for the possibility that the observed
effect is partially or entirely due to systematic error rather than a true
causal effect.

**Interpretation guide:**

| Scenario | Nominal p | Calibrated p | Interpretation |
|---|---|---|---|
| No bias | 0.003 | 0.005 | Minimal calibration adjustment; result is robust |
| Moderate bias | 0.003 | 0.08 | Systematic error inflates significance; interpret cautiously |
| Severe bias | 0.003 | 0.45 | Result is likely driven by bias, not a true causal effect |


## Integration with Diagnostics

Negative controls integrate into the broader Medusa diagnostic framework
through `runInstrumentDiagnostics()`. When `nc_outcome_*` columns are
present in the cohort data, the function automatically runs the negative
control analysis and includes the results in its output:

```{r diagnostics-integration, eval = FALSE}
diagnostics <- runInstrumentDiagnostics(
  cohortData = cohortData,
  covariateData = covariateData,
  instrumentTable = instruments
)

# NC results are included in the diagnostics object
diagnostics$negativeControlResults
diagnostics$diagnosticFlags$negativeControlFailure
```

The `diagnosticFlags$negativeControlFailure` flag is TRUE when the
negative-control analysis detects aggregate systematic bias in the null
distribution, rather than when any single control happens to cross a
nominal p-value threshold. This flag appears alongside other diagnostic
checks (weak instruments, allele frequency discrepancies, genotype
missingness) in the Medusa results dashboard.


## The Shiny Results Explorer

The interactive results explorer (`launchResultsExplorer()`) includes a
dedicated Negative Controls tab when negative control results are
provided. This tab displays:

- **Summary cards** showing the number of NC outcomes tested, whether
  bias was detected, and the calibrated primary p-value when available
- **Forest plot** of NC effect estimates with confidence intervals
- **Calibrated primary estimate** (if available)
- **Table** of all per-outcome results for detailed inspection

```{r shiny-explorer, eval = FALSE}
launchResultsExplorer(
  mrEstimate = estimate,
  sensitivityResults = sensitivity,
  diagnosticResults = diagnostics,
  instrumentTable = instruments,
  negativeControlResults = ncResults
)
```


## Worked Example: Full Pipeline

Putting it all together with simulated data:

```{r full-pipeline}
# 1. Simulate data
simData <- simulateMRData(n = 3000, nSnps = 5, trueEffect = 0.5, seed = 99)

# 2. Add negative control outcomes (true effect = 0 for all)
cohortData <- simulateNegativeControlOutcomes(
  simData$data,
  nControls = 20,
  seed = 100
)

# 3. Run the negative control analysis
ncResults <- suppressMessages(
  runNegativeControlAnalysis(
    cohortData = cohortData,
    instrumentTable = simData$instrumentTable
  )
)

# 4. Summarize
cat("Bias detected:", ncResults$biasDetected, "\n")
cat("Mean NC beta_MR:", round(mean(ncResults$ncEstimates$beta_MR, na.rm = TRUE), 4), "\n")
cat("NC outcomes analyzed:", nrow(ncResults$ncEstimates), "\n")

# 5. Look at the estimates
head(ncResults$ncEstimates[, c("outcome_id", "beta_MR", "se_MR", "pval")], 10)
```

Because the simulated negative control outcomes are independent of
genotype, the MR estimates should be small and centered near zero, and
`biasDetected` should be FALSE. This confirms that the analysis pipeline
is not introducing systematic bias for this dataset.


## When Things Go Wrong

If `biasDetected` is TRUE or many negative controls are significant,
consider:

1. **Population stratification.** Are the genetic instruments correlated
   with ancestry or geography? This can create spurious associations with
   many outcomes simultaneously.
2. **Horizontal pleiotropy.** If the instruments affect many traits beyond
   the exposure, negative controls that share biological pathways with
   those traits may show non-null effects.
3. **Collider bias.** If the cohort is selected on a variable influenced
   by both the instruments and the outcomes, negative controls can show
   spurious associations.
4. **Technical artifacts.** Genotype missingness, allele frequency
   discrepancies, or coding errors can create systematic bias across all
   outcomes.

In any case, use the calibrated p-value rather than the nominal p-value
for the primary analysis, and report both values transparently.


## References

- Schuemie MJ, Ryan PB, DuMouchel W, Suchard MA, Madigan D (2014).
  Interpreting observational studies: why empirical calibration is needed
  to correct p-values. *Statistics in Medicine*, 33(2), 209--218.
- Schuemie MJ, Hripcsak G, Ryan PB, Madigan D, Suchard MA (2018).
  Empirical confidence interval calibration for population-level effect
  estimation studies in observational healthcare data. *PNAS*, 115(11),
  2571--2577.
- Davies NM, Holmes MV, Davey Smith G (2018). Reading Mendelian
  randomisation studies: a guide, glossary, and checklist for clinicians.
  *BMJ*, 362, k601.
- Sanderson E, Glymour MM, Holmes MV, Kang H, Morrison J, Munafo MR,
  Palmer T, Schooling CM, Wallace C, Zhao Q, Davey Smith G (2022).
  Mendelian randomization. *Nature Reviews Methods Primers*, 2, 6.
